
\section{Results from Prior NSF Support}
\label{ssec:prior}

\paragraph{Expedition in Computing: The Science of Deep Specification (PI: Zhong Shao)} 
NSF CCF-1521531, \$2,046,445, with Andrew Appel and Lennart Beringer
(Princeton), Benjamin Pierce and Stephanie Weirich and Steven
Zdancewic (U. Pennsylvania), and Adam Chlipala (MIT), 2015-2020.  {\em
  Intellectual Merit:} The Yale Component of this expedition project
aims to develop a concurrent certified OS kernel (CertiKOS) and
connect it with the verified RISC-V hardware (developed at MIT) and
the web server (developed at UPenn, and verified at Princeton). The
key emphasis is to work out the detailed specification for the machine
interfaces (e.g., for RISC-V) and the system call interface (e.g., for
CertiKOS) so that software and hardware components verified by
multiple DeepSpec groups can indeed be linked togher. During the first
two years, The Yale team has successfully developed a clean-slate
CertiKOS hypervisor OS kernel that runs successfully on both Intel and
AMD multicore platforms with hardware virtualization and can boot
Ubuntu or Debian Linux as guest~\cite{certikos16}.  We have also
developed a new compositional approach for formally specifying and
verifying sequential and concurrent OS
kernels~\cite{chen16,costanzo16,certikos16}.  {\em Broader Impacts:}
This award has partly supported multiple postdocs and students in the
past two years. PhD student Ronghui Gu will join Columbia CS as a new
tenure track Assistant Professor.  The team has organized multiple
outreach workshops in 2016-2017, and also a two-week DeepSpec summer
school in 2017 with more than 150 attendees from all over the
world. PI Shao has incorporated the layered CertiKOS kernel into an
innovative undergraduate OS class.  {\em Representative Publications}:
two PLDI papers~\cite{chen16,costanzo16} and one OSDI
paper~\cite{certikos16}.

\paragraph{Co-PI Jeremie Koenig}
There is no prior NSF support for which Koenig is a PI.

