\documentclass[11pt]{article}
\usepackage{times}

\usepackage{ifthen}
\usepackage{latexsym}
\usepackage{url}
\usepackage{listings} %for code
%\usepackage{mathpartir} %for typing rules

\usepackage{alltt}
\usepackage{multirow}
\usepackage{paralist}
\usepackage{graphicx} %to include images
\usepackage{scalerel}
\usepackage{relsize}

\usepackage{wrapfig}

%\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amsthm}
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings, multicol} %for code
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{bussproofs} %for proof trees
\usepackage{comment}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, decorations.pathmorphing}
\usepackage{paralist}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{relsize}
\usepackage[font=small,labelfont=bf]{caption}\usepackage[normalem]{ulem}
\usepackage{cancel}
\usepackage{caption}
\usepackage{lipsum}
\usepackage{xcolor}

\newcommand{\DeepSpec}{\texttt{DeepSEA}\xspace}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}


\long\def\cruzica#1{{\color{red}{\bf Ruzica: }{\small [#1]}}}
\sloppypar

%\include{defs}
\input{local}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\voffset             0in    %  top vertical offset
\hoffset             0in    %  left horizontal offset
\oddsidemargin       0pt    %  Left margin on odd-numbered pages.
\evensidemargin      0pt    %  Left margin on even-numbered pages.
\topmargin           0pt    %  Nominal distance from top of page to top of
\headheight          0pt    %  Height of box containing running head.
\headsep             0pt    %  Space between running head and text.
\textwidth         6.5in    %  Width of text on page
\textheight          9in    %  Height of text on page
%\setlength{\parskip}{.035in}
\renewcommand{\floatpagefraction}{.9}
\renewcommand{\textfraction}{0.1}
%\renewcommand{\baselinestretch}{1.03}

\let\cmttdfl\ttdefault          %what for?
\let\cmsfdfl\sfdefault          %what for?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\setcounter{page}{1}
\pagenumbering{roman}

%\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Cover Page} 

%\begin{comment}
\vspace*{1in} 
\centerline{
\begin{tabular}{c}
  {\Large SHF: Medium: End-to-End and Compositional Verified Secure Compilation}
\\[10ex] 
\large Zhong Shao (PI) \\[.5ex]
\large Jeremie Koenig (co-PI) \\[.5ex]
\large Department of Computer Science \\[.1ex] 
\large Yale University \\[.1ex]
\large 51 Prospect Street \\[.1ex]
\large New Haven, CT 06520-8285, USA \\[.5ex]
\large Phone: 203-432-6828 \\[.1ex]
  \large Email: \{zhong.shao, jeremie.koenig\}@yale.edu \\[5ex]
%%
  \large {\em{} A proposal for CISE CCF Core Program on Software and Hardware Foundations (SHF)}\\
\large {\em{} NSF program solicitation 21-616}
\end{tabular}
}
\vspace*{1in} 
\newpage
%\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{Project Summary}

\thispagestyle{empty}
\input {summary}
\newpage


\pagenumbering{arabic}
\setcounter{page}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Project Description}

\section{Introduction}
\label{sec:intro}


Distributed systems are difficult to reason about and verify because
they employ sophisticated protocols to coordinate failure-prone
distributed nodes over an unreliable network.  Their correctness often
relies on subtle, implicit global invariants, which hinder the ability
to understand why the system works and how one system relates to
another. To further complicate matters, modern applications build on
top of multiple distributed systems~\cite{hadoop, hadoopeco}, which
necessitates reasoning about concurrent interactions between systems.


Formal models and verification of the code are necessary to guarantee
the safety and correctness of a distributed system. However, it is
challenging to verify a standalone distributed system and harder still
to verify multiple interacting distributed
systems~\cite{shimlayer}. There exist amazing efforts to apply formal
reasoning and verification to distributed systems that provide useful
tools for simplifying some aspects of verification, such as high-level
state machine refinement methodology~\cite{ironfleet}, network model
transformation~\cite{verdi}, automatic invariant generation~\cite{i4},
and modular reasoning for multiple distributed protocols with simple
interactions~\cite{disel}. However, they work at a relatively low
level while dealing with implementation details, so verifying
individual distributed systems requires redundant
reasoning. Similarly, the low-level details make it difficult to
address the new challenges that come with composition of systems.

Then how can we facilitate the verification of multiple individual
distributed systems and their compositions? We claim that providing a
simple, atomic, and general interface is key to reducing the
complexity of verifying distributed systems. It is well known that
providing a simple object model by abstracting away implementation
details can greatly ease the understanding and verification of complex
systems~\cite{herlihy90, herlihy08book, wormspace, concurrency}. While
existing efforts listed above do employ some amount of abstraction,
non-essential details, such as message orderings and quorum sizes,
sometimes leak through into the high-level reasoning.  Thus, by
modeling distributed systems at an even higher level, one can focus on
just the core logic, and deal with the implementation details
separately.

We propose to design a new high-level abstraction of distributed
systems to justify our claim. We aim to realize a specification and
verification framework, \sysname{}, for individual distributed systems
and their composition surrounding the abstraction, and model and
verify distributed systems down to the code level.  \sysname{} stands
for \sysnamelong{} and targets a general verification approach for
strongly consistent replicated distributed systems that satisfy the
replicated state safety property (see
Theorem~\ref{thm:atomic-spec-prefix}).  Examples of such systems
include multi-Paxos~\cite{rvrpaxos}, Raft~\cite{raft}, and
chain-replication~\cite{chainreplication}, which are widely used in
production systems.  \sysname{} aims to model and verify composition
of such systems as long as the composition satisfies the replicated
safety property.

Inspired by work on shared-memory concurrent
objects~\cite{herlihy08book, lintextbook}, we propose a novel model of
distributed systems called the atomic distributed object (ADO) and
plan to build \sysname{} surrounding the ADO.  The ADO models all
distributed operations as atomic functions on a non-distributed shared
object, but with added non-determinism to reflect its distributed
nature. The calls are captured by atomic updates to a linear, logical
history of events. The ADO model is detailed enough to express
high-level behaviors of distributed systems, but abstract enough that
implementation details do not unnecessarily complicate the high-level
reasoning.

We plan to divide \sysname{} into four major components
(Figure~\ref{fig:advert}) and implement each component in
Coq~\cite{coq}: 1) the ADO specification, 2) a multi-ADO reasoning
framework, 3) a network-based specification, and 4) a code-level
verification framework.  The ADO specification sits at the center of
\sysname{} to model individual distributed systems using a simple
atomic interface. The multi-ADO reasoning framework allows composite
systems to be built from ADOs and reasoned about modularly. \sysname{}
fills in the missing details from below with the network-based
specification, and provides a clear path between implementation and
ADO by proving a refinement relation. This specification will be
parameterized such that it can model most protocols with similar
network patterns. Along the way we will solve the challenges that stem
from mapping arbitrarily delayed future events in the network-level
specification to an atomic function call in the ADO model. Verifying
that different systems refine the ADO model will be made simple and
convenient through the network-based specification because the
properties captured by the common network pattern could allow for the
reuse of proofs. Finally, the code level verification framework will
use the certified concurrent abstraction layer
(CCAL)~\cite{concurrency} approach (the PI's prior work).  \sysname{}
will take care of each step to connect the code-level implementation
of a system up to multiple distributed system reasoning.

Our proposed research consists of the following four components:
\begin{itemize}%[leftmargin=*]
\item We will identify and propose an atomic distributed object model
  that preserves the key characteristics of distributed systems, but
  hides the implementation details.
\item We will explore different distributed protocols to create
  parameterized network-based specifications that connect individual
  system implementations to the atomic distributed object model
  through a refinement relation. Each network-based specification will
  reuse the refinement proofs and act as a template that connects a
  similar class of individual systems to the atomic distributed object
  model.
\item We will conduct studies of composite distributed applications to
  illustrate how \sysname{} and atomic distributed object model enable
  simple reasoning about multiple distributed system interactions. We
  will explore various patterns of system compositions for broad
  impact.
\item To demonstrate the power and real-world applicability of
  \sysname{}, we will carry out concrete instantiations multiple
  distributed protocols, including multi-Paxos, and Raft, and extend
  the instantiations to a large-scale distributed shared memory
  system.
\end{itemize}

\paragraph{Related Efforts} The PI is a co-investigator of the NSF-funded DeepSpec/CertiKOS project. 
The \sysname{} project differs from the CertiKOS project as it focuses
on distributed systems which run above operating system
layers. CertiKOS tackles challenges of verifying operating system code
base and handling concurrency within operating system. The \sysname{}
project is different from the CertiKOS project as it reasons about
distributed system concurrency, where completely different failure and
network conditions are present. The \sysname{} project rather
complements the CertiKOS project by extending the verified software
stack up to a composition of distributed applications.

Other efforts to facilitate the verification of distributed systems
are introduced at the start of this section~\cite{ironfleet, verdi,
  i4, disel}. All these efforts do little to explore how to expose the
common high-level behaviors of distributed systems and hide the
implementation details. They do not explore reusable proofs to verify
multiple distinct distributed systems using a common framework and do
not focus on verifying compositions of multiple distributed
systems. Our grand goal is to realize a distributed system modeling
and verification framework that can verify both standalone and
composite distributed systems to create a fully verified distributed
ecosystem.


\section{Compostional Verified Compilation}

\section{End-to-End Verified Compilation}

\section{CompCertOX and Support for CCAL}
    
\section{Verified Secure Compilation}


\section{Intellectual Merit}

Existing distributed system verification efforts have been focused
mostly towards handling complex low-level details of individual
distributed systems, whereas little attention has been paid towards
verifying multiple individual distributed systems and their
interactions. The proposed research highlights this overlooked
challenges and proposes a novel verification and reasoning approach
surrounding an atomic distributed object (ADO).  Our research will
introduce two new capabilities that have not been explored: 2) the ADO
will trivialize the verification of multiple distributed system
interactions by hiding the low-level details of the system while
exposing enough semantics to reason about the system compositions; and
2) \sysname{} will significantly simplify the verification of
individual systems with a new approach of capturing common distributed
system properties and exposing reusable proofs structures through
network-based specifications.

Our research aims to study and verify as many well-established
distributed systems as possible to extract the generic properties of
the systems and common patterns of their compositions. Our goal is to
establish concrete models and verification frameworks for different
levels of abstractions and apply the findings to verify conventional
and cutting-edge distributed systems down to the code level. Similar
to how ADO explains the distributed system behavior at a high-level,
each step of our research will cover low to high level details of
distributed systems and promote the general understanding of complex
distributed systems. We expect the research outcome will greatly
contribute to guaranteeing the safety and correctness of large-scale
systems and applications that are widely used everyday for example
through the cloud.

\section{Broader Impact}

Our proposed research will contribute greatly to build a safe and
bug-free distributed software ecosystem. Especially, the parameterized
verification framework and atomic-distributed-object-based composite
reasoning of distributed systems capture the core properties of
distributed systems. Because most seemingly different distributed
systems are developed and optimized while preserving these core
properties, our verification approach will be able to host a wide
range of distributed systems and tolerate frequent system updates to
easily verify the safety properties. In addition, our verification
framework will guarantee the safety of large-scale distributed systems
that were considered very challenging. We expect that our research
will considerably lighten the burden of distributed system testing and
debugging by making distributed system verification greatly
approachable and applicable.

Furthermore, our research can be easily extended to different domains,
such as cyber physical systems and internet of things, where hundreds
of devices realize a form of a distributed system. For example,
mission critical health monitoring wireless sensor nodes and
coordination of self-driving cars and avionics would be able to
directly benefit from our research. As computing devices are becoming
more and more common, our research will be able to verify the
correctness of their interactions and have a greater impact.

We plan to develop innovative distributed system software and formal
verification curriculum at both the undergraduate and graduate
level. PI Shao recently started to teach a course on language-based
security at Yale and the term projects are designed based on initial
findings of our proposal.  Extending the term project, co-PI Shin is
planning to develop a distributed system verification course, where
students can gain hands-on experience on both system design and
verification. We plan to extend the courses to undergraduate and
Master's research program where students can participate in the
proposed research. We will make our course materials freely available
and encourage our colleagues to use them at other universities.

We are planning to further use our research materials to foster
diversity and inclusion. Co-PI Soul\'e has been specially focusing
on supporting Latin American and Hispanic students for few years with
scholarship programs, student visit programs, and diversity seminars. 
The PIs plan to continue these efforts jointly and to recruit 
underrepresented students to get involved in our research.

Several composite distributed system examples that we plan to model
and verify are motivated by experiences in the commercial sector, 
and therefore has a clear promise for industrial impact. 
For example, co-PI Soul\'{e} has worked closely with Western Digital 
on a project that uses network hardware to build a new breed of 
distributed shared memory system. Successful modeling and
verification of the system can directly influence the safety and 
correctness of the production prototype. 

The design, implementation, and validation of a software artifact 
is one of our great contributions to our proposal. 
We plan to release the software developed in this project under 
a flexible, open-source license.


\section{Results from Prior NSF Support}
\label{ssec:prior}

\paragraph{Expedition in Computing: The Science of Deep Specification (PI: Zhong Shao)} 
NSF CCF-1521531, \$2,046,445, with Andrew Appel and Lennart Beringer
(Princeton), Benjamin Pierce and Stephanie Weirich and Steven
Zdancewic (U. Pennsylvania), and Adam Chlipala (MIT), 2015-2020.  {\em
  Intellectual Merit:} The Yale Component of this expedition project
aims to develop a concurrent certified OS kernel (CertiKOS) and
connect it with the verified RISC-V hardware (developed at MIT) and
the web server (developed at UPenn, and verified at Princeton). The
key emphasis is to work out the detailed specification for the machine
interfaces (e.g., for RISC-V) and the system call interface (e.g., for
CertiKOS) so that software and hardware components verified by
multiple DeepSpec groups can indeed be linked togher. During the first
two years, The Yale team has successfully developed a clean-slate
CertiKOS hypervisor OS kernel that runs successfully on both Intel and
AMD multicore platforms with hardware virtualization and can boot
Ubuntu or Debian Linux as guest~\cite{certikos16}.  We have also
developed a new compositional approach for formally specifying and
verifying sequential and concurrent OS
kernels~\cite{chen16,costanzo16,certikos16}.  {\em Broader Impacts:}
This award has partly supported multiple postdocs and students in the
past two years. PhD student Ronghui Gu will join Columbia CS as a new
tenure track Assistant Professor.  The team has organized multiple
outreach workshops in 2016-2017, and also a two-week DeepSpec summer
school in 2017 with more than 150 attendees from all over the
world. PI Shao has incorporated the layered CertiKOS kernel into an
innovative undergraduate OS class.  {\em Representative Publications}:
two PLDI papers~\cite{chen16,costanzo16} and one OSDI
paper~\cite{certikos16}.

\paragraph{Co-PI Jeremie Koenig}
There is no prior NSF support for which Koenig is a PI.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{References Cited}
{\bibliographystyle{abbrv}
\makeatletter
\renewcommand\small{%
   \@setfontsize\small{10.46}{12.77}
   \abovedisplayskip 10\p@ \@plus2\p@ \@minus5\p@
   \abovedisplayshortskip \z@ \@plus3\p@
   \belowdisplayshortskip 6\p@ \@plus3\p@ \@minus3\p@
   \def\@listi{\leftmargin\leftmargini
               \topsep 6\p@ \@plus2\p@ \@minus2\p@
               \parsep 3\p@ \@plus2\p@ \@minus\p@
               \itemsep \parsep}%
   \belowdisplayskip \abovedisplayskip
}
\makeatother
 \small\bibliography{shao}
} \newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

