\documentclass{beamer}
\usepackage{stmaryrd}
\usepackage{ebproof}

\begin{document}

\begin{frame}{Full abstraction and semantics preservation}
Two commonly used notions of compiler correctness:
\begin{itemize}
  \item \textbf{Semantics preservation}
  \[
    \begin{prooftree}
      \hypo{C(P) = P'}
      \infer1{\llbracket P \rrbracket_S \le \llbracket P' \rrbracket_T}
    \end{prooftree}
  \]
  This is a strong property but depending on the specifics, \\
  it does not necessarily entail preservation of security.
  \item \textbf{Full abstraction}
  \[
    \begin{prooftree}
      \hypo{C(P_1) = P_1'}
      \hypo{C(P_2) = P_2'}
      \infer2{
        \llbracket P_1 \rrbracket_S \simeq \llbracket P_2 \rrbracket_S
        \: \Leftrightarrow \:
        \llbracket P_1' \rrbracket_T \simeq \llbracket P_2' \rrbracket_T
      }
    \end{prooftree}
  \]
  This does not actually relate source and target behaviors, \\
  but is suitable for preservation of noninterference, etc.
\end{itemize}
\end{frame}

\begin{frame}{Security preservation and refinement}
Consider:
\begin{itemize}
  \item
    Semantics preservation and fully abstract compilation \\
    are incomparable.
  \item
    In principle,
    refinement ($\le$) is all about preserving properties:
    \[
      x \le y \: \Longleftrightarrow \:
      \forall P \in \mathbf{Prop} \mathbin. P(x) \Rightarrow P(y)
    \]
  \item Yet fully abstract compilation is how you would typically \\
    preserve security properties like noninterference.
\end{itemize}

\vfill
How can this be reconciled?
\begin{itemize}
  \item $\mathbf{Prop}$ may not include the security properties we want.
  \item Full abstraction is sufficient but not necessary.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Nondeterminism}
Nondeterminism is at the root of the discrepancy.

\vfill
For example the seemingly secure specifications:
\begin{verbatim}
int secret_flag;
int foo() { return any_coin() ? 0 : 1; }
\end{verbatim}
or
\begin{verbatim}
int secret_flag;
int foo() { *NULL; }
\end{verbatim}
are refined by the obviously insecure:
\begin{verbatim}
int secret_flag;
int foo() { return secret_flag ? 0 : 1; }
\end{verbatim}
\end{frame}

\begin{frame}[fragile]{The role of calling conventions}
Suppose
\begin{verbatim}
  int secret;
  int zero() { return 0; }
\end{verbatim}
compiles to:
\begin{verbatim} 
  zero:
    mov secret, %ecx
    mov $0, %eax
    ret
\end{verbatim}
The C source program is secure but can be correctly compiled \\
to seemingly insecure target assembly code.
\end{frame}

\end{document}
