\section{Concurrency}
\label{sec:concurrency}

While taking concurrency into account is necessary
to model most real-world heterogeneous systems,
doing so introduces much complexity, and
models with concurrency are usually
harder to mechanize and reason about.
For example, in sequential programs,
control follows a strict alternation
between the system being modeled and its environment;
this constraint is lost in the presence of concurrency.
The uncertainty associated with scheduling also
introduces a high degree of nondeterminism.

\subsection{Approach}

To manage this complexity,
we will follow a similar approach as we did in \S\ref{sec:effects},
and decouple concurrent layer specifications from
the sequential, deterministic per-thread code that invokes them.
In addition,
using \emph{linearizability} as a correctness criterion
allows us to formulate concurrent specifications in forms that
\begin{itemize}
  \item preserve a strict alternation between
    the system and the environment (atomicity);
  \item factors out scheduling nondeterminism
    into the correctness criterion itself.
\end{itemize}
As such,
the simplest version of our concurrent model
will use a representation for
layer interfaces and implementations
that remains very close to the sequential version,
facilitating the incorporation of refinement conventions and
making sure straightforward specification and verification tasks
can be carried out without undue complexity.

At the same time,
not every concurrent object can be specified
using a deterministic, atomic linearized specification.
To deal with more sophisticated situations,
we will build on the techniques discussed in \S\ref{sec:effects}
to generalize our model by allowing
nondeterministic and non-atomic specifications.

\subsection{Atomic Linearized Specifications}

When multiple threads interact with a concurrent object,
their calls into the object can overlap in unrestricted ways.
The object's behavior in one thread can then be
influenced by any concurrent invocations happening in other threads
while the method is executed.
However,
many concurrent objects behave \emph{as if}
their methods were executed atomically,
at some point between their observable invocation and return.
When that is the case,
a concurrent object can be described
using a specification that remains
essentially sequential and deterministic.
In that case,
the concurrent object implementation
is said to \emph{linearize} to
the sequential atomic specification.

Linearized atomic specifications are
especially useful in our case
as they allow us to avoid much of the complexity associated with
describing concurrent behaviors.
A specification of this kind is essentially
a sequential layer specification as described in \S\ref{sec:effects}
operating over a signature
\newcommand\parr{!}
\[
  \parr E := \{ \alpha \mathbin. m \mathbin: N \mid
           \alpha \in \varGamma \wedge
           (m \mathbin: N) \in E \}
  \,,
\]
where each operation is annotated with
a thread identifier $\alpha$
taken from a predefined, fixed set of thread names $\varGamma$.
In this form,
linearized specifications are much easier to
incorporate in a framework with effects and refinement conventions,
which are inherently sequential in many aspects.

\subsection{Refinement Framework}

With this in mind,
following the strategy established in \S\ref{sec:effects},
the concurrent version of our refinement framework
will be constructed using:
\begin{itemize}
  \item linearized atomic layer specifications as described above;
  \item the sequential (per-thread) implementation strategies
    found in our base framework;
  \item a notion of layer correctness described below.
\end{itemize}
We will write
$\Gamma \vdash_\mathbf{R} \sigma : \Delta$
when a per-thread strategy
$\sigma : E \twoheadrightarrow F$,
interacting with an underlay concurrent specification
$\Gamma : M \looparrowright {!E}$,
implements an object which linearizes to
$\Delta : M \looparrowright {!F}$
\emph{up to} the refinement convention $\mathbf{R}$.
Conceptually,
this situation can be depicted as
\begin{equation}
  \Gamma \vdash_\mathbf{R} \sigma : \Delta
  \qquad \qquad
  \begin{tikzcd}
    M \ar[d, equal] \ar[rr, "\Delta"] & &
    F_1 \ar[d, "\mathbf{R}", leftrightarrow] \\
    M \ar[r, "\Gamma"] &
    E \ar[r, "\sigma", twoheadrightarrow] & F_2
  \end{tikzcd}
  \label{eqn:concref}
\end{equation}
However,
note that in general
the composite $\sigma[Gamma]$
cannot be defined,
as a \emph{most general linearized atomic specification}
implemented by $\sigma$ may not exist.
In other words,
the property described by (\ref{eqn:concref})
should be seen as a monolithic correctness property
linking the four components $\Delta$, $\Gamma$, $\sigma$ and $\mathbf{R}$.

Nevertheless,
the visualization provided above
is useful to understand how
concurrent refinements
interact with our sequential implementations framework.



\subsection{Effects}

\subsection{Liveness}

\subsection{Shared Memory}

Mention memory PCM and angelic CompCert sem?

\subsection{Non-atomic Objects}

\paragraph{Task 2a: LinCCAL-style correctness with atomic linearized specs}

\paragraph{Task 2b: Liveness properties}

\paragraph{Task 2c: Angelic shared memory}

\paragraph{Task 2d: Non-atomic behaviors}

