%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% yale_thesis.tex
% Alexander Cerjan
% 2014/04/07
%
% A bare, sample template for a Yale PhD thesis using yalephd.cls
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[letterpaper,11pt]{yalephd}
% remove draft option for final printing.
% font size must be between 10pt-12pt.

\usepackage{geometry} % you need this for yalephd.cls to work.
\usepackage{graphicx} % you probably want the rest of these.
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{appendix}
\usepackage{comment}
\usepackage{cmll}
% \usepackage{cite}
% \usepackage{notoccite}

\usepackage{ebproof}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{subcaption}

\usepackage[square,sort,semicolon]{natbib}
%\usepackage[draft=false,allcolors=blue,colorlinks]{hyperref}
\usepackage[hidelinks]{hyperref}

\usepackage{minted}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{booktabs}

\usetikzlibrary{arrows.meta,positioning,fit,calc}

\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{challenge}{Challenge}

\hyphenation{Comp-Cert}
\hyphenation{Comp-CertX}
\hyphenation{Comp-CertO}
\hyphenation{Comp-CertOX}
\hyphenation{Comp-CertOE}
\hyphenation{Comp-CertM}
\hyphenation{Certi-KOS}

% \bibliographystyle{abbrvunsrt}

\begin{document}

% Need to define title before the abstract.
\title{Building Certified Systems with Compositional Semantics}
% capitalize the first letter in every word except conjunctions and prepositions
\author{Yu Zhang}
\advisor{Zhong Shao}
\date{December, 2025} % usually not \today.

\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\kw}[1]{\ensuremath{ \mathsf{#1} }}
\newcommand{\C}{\ensuremath{ \mathcal{C} }}
\newcommand{\A}{\ensuremath{ \mathcal{A} }}
\newcommand{\que}{\circ}         % superscript for questions
\newcommand{\ans}{\bullet}       % superscript for answers
\newcommand{\termi}[1]{\langle {#1} ]}
\newcommand{\encap}[1]{[ {#1} \rangle}
\newcommand{\at}{\mathbin@}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\ifr}[1]{\mathrel{[{#1}]}}
\newcommand{\idsc}{\mathbf{id}} % identity simulation convention
\newcommand{\intl}[1]{#1^0}
\newcommand{\envstep}{\leadsto}
\newcommand{\sysstep}{\rightarrowtail}
\newcommand{\emptysig}{0}
\newcommand{\vcomp}{\fatsemi}
\newcommand{\jr}{\mathsf{Y}}
\newcommand*{\twoheadleftrightarrow}{%
  \twoheadleftarrow
  \mathrel{\mkern-15mu}%
  \twoheadrightarrow
}
\newcommand{\lensarrow}{\rightleftarrows}

% String diagrams {{{
\colorlet{scsdbg}{lightgray!50!white}
\colorlet{tssdbg}{lightgray!50!white}
\colorlet{memsdbg}{blue!50!white}
\colorlet{mmemsdbg}{blue!50!white}
\colorlet{penvsdbg}{green!50!white}
% }}}

% All the stuff at the front of your thesis.
\frontmatter

\begin{abstract}

  Formal verification has emerged as
  the most rigorous approach for building reliable computer systems.
  Certified systems, in particular,
  are accompanied by a formal specification
  and a machine-checkable proof of correctness,
  enabling independent verification by third parties.

  However, verifying large-scale, heterogeneous systems
  presents significant challenges that current techniques cannot adequately address.
  Such systems require compositional methods
  that can flexibly integrate diverse verification approaches,
  including program logic, certified compilation, and data abstraction.
  While compositional semantics offers a promising foundation,
  existing frameworks are typically built on specialized operational models
  that resist integration with one another.

  This dissertation presents a three-dimensional algebraic compositional structure
  that unifies diverse verification techniques within a single coherent foundation
  and enables systematic composition across program modules,
  abstraction levels,
  and system state components.
  This work first extends CompCertO to create the OpenTX and OpenTE frameworks,
  which conform to the three-dimensional compositional structure
  and demonstrate practical verification capabilities.
  Building on these foundations, the dissertation then develops
  a more general strategy-based semantic model with refinement conventions
  that also conforms to the three-dimensional approach
  and provides the core framework for compositional verification.
  All components are fully mechanized in the Rocq proof assistant,
  and their capabilities are demonstrated
  through multiple case studies.

\end{abstract}

\maketitle
\makecopyright{2025} % change as needed.
\tableofcontents
\listoffigures % remove this if you have no figures.
\listoftables % remove this if you have no tables.

\chapter{Acknowledgements} % this needs to be before \mainmatter.

I am deeply grateful to my advisor,
Prof. Zhong Shao, for his invaluable guidance,
encouragement,
and unwavering support throughout my PhD journey.
During challenging times---especially throughout the pandemic and
while navigating VISA issues---he stood by me,
offering both understanding and
practical help so that I could remain focused on my research.

I owe special thanks to Jérémie Koenig,
with whom I had the great fortune to work closely
for countless hours.
Together,
we explored sophisticated semantic models,
developed Rocq proofs,
and co-authored papers.
His generosity in sharing knowledge,
his patience in explanation,
and his enthusiasm in research
have been a constant source of inspiration.
Many of the core ideas in this dissertation
originate from his insights,
for which I am sincerely indebted.

I would also like to thank
Profs. Ruzica Piskac, Yuting Wang, and Alex Lew
for serving on my thesis committee.
Their thoughtful feedback has greatly improved this work.
I am particularly grateful to Prof. Yuting Wang
and his student Ling Zhang,
whose discussions have been especially helpful.

I am fortunate to have shared my time
at Yale
with the supportive community of the FLINT group
and the broader CS department.
I would like to especially acknowledge
Yuyang Sang, Vilhelm Sjöberg, Bowen Huang, Jialu Zhang,
Ning Luo, Arthur Oliveira Vale, and Yixuan Chen,
along with many others,
for their collaboration, advice, and friendship.

Finally,
I owe my deepest gratitude to my parents,
whose unwavering love and support
have been the foundation of all my pursuits.
Their encouragement has given me
the strength to follow my path with confidence.

% Starts proper arabic numbering of pages and chapters.
\mainmatter

\input{introduction}

\input{background}

\input{compcertox}

\input{compcertoe}

\input{strategy}

\input{refconv}

\input{application}

\input{related}

\input{conclusion}

% Any chapters such as End Notes go after this.
\backmatter

% \bibliography{references}
% for your own sake, use a bibtex file, so all of the numbering of references will be done
% automatically.

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\bibname}
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
